---
title: "Analysis on Hate Crime Rates Across the U.S. (Part 2)"
author: "Chloe Song"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
  word_document: default
fontsize: 11pt
geometry: margin=1in
---
GitHub Repository: https://github.com/ChloeSong2468/Study-on-Hate-Crime-Rates-across-United-States

```{r setup, include=FALSE}
#load  packages
library(tidyverse)
```

```{r load_data, include=FALSE}
#source: https://github.com/fivethirtyeight/data/blob/master/hate-crimes/hate_crimes.csv
#load dataset
hate_crimes <- read_csv("hate_crimes.csv")

#remove rows missing the SPLC hate crime rate, add abbrev
hc <- hate_crimes %>%
  filter(!is.na(hate_crimes_per_100k_splc)) %>%
  mutate(
    state_abbrev = state.abb[match(state, state.name)],
    state_abbrev = ifelse(state == "District of Columbia", "DC", state_abbrev)
  )
```

# Introduction

Hate crimes are a serious concern across the United States, and understanding what
socioeconomic factors are linked to them can help inform our policies. This project
looks into whether states with higher median household incomes have different hate crime
rates than states with lower median household incomes. The research question is: 

“Is the distribution of hate crime rates different between states with above-median household income and states with below-median household income?”

In Part 1, the EDA showed that the hate crime rate is right-skewed across the
states, with some high rate outliers pulling the mean above the median. Somewhat
surprisingly, the correlation between median household income and hate crime rate was weakly
positive (r = 0.351), suggesting that states with higher income do not tend to have lower hate
crime rates. This report tests whether that difference is statistically significant.

# Methods

## Test Selection

To answer the research question, we use a Wilcoxon rank-sum test. States are split 
into two groups based on whether their median household income is above or below the 
median income across all states. The test compares the distribution of hate crime rates 
between these two groups.

```{r group_creation, echo = FALSE}
#compute the median household income across all states
median_income <- median(hc$median_household_income, na.rm = TRUE)

#make income group variable: above or below median income
hc <- hc %>%
  mutate(income_group = ifelse(median_household_income >= median_income,
                               "Above Median", "Below Median"))

#check group sizes
table(hc$income_group)
```

The Wilcoxon rank-sum test is appropriate here because:
1. We are comparing a continuous outcome (hate crime rate) across two independent groups
(above vs. below median income).
2. The two groups are independent; a state's income group has no influence on another state's.
3. The test assesses whether one group tends to have systematically higher or lower hate crime
rates than the other.


## Hypotheses

$F_1$ = distribution of hate crime rates for above-median income states
$F_2$ = distribution of hate crime rates for below-median income states

$$H_0: F_1 = F_2$$
$$H_a: F_1 \neq F_2$$
We use a significance level of $\alpha = 0.05$.


## Assumption Verification

**1. Independence:** Each observation is a U.S. state. States are independent units; the
hate crime rate in one state does not directly influence the rate in another.

**2. Continuity:** The hate crime rate is a continuous variable, satisfying this requirement
for the Wilcoxon rank-sum test.

**3. Similar shape:** The Wilcoxon rank-sum test assumes the two distributions have a roughly
similar shape. We verify this with density plots.

```{r densityplot, fig.height=3.5, echo = FALSE}
# Density plots to compare the distribution shape of each group
ggplot(hc, aes(x = hate_crimes_per_100k_splc, fill = income_group)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("Above Median" = "dodgerblue2", "Below Median" = "tomato3")) +
  labs(
    title = "Distribution of Hate Crime Rate by Income Group",
    x     = "Hate Crimes per 100,000 (SPLC)",
    y     = "Density",
    fill  = "Income Group"
  ) +
  theme_minimal()
```

Both distributions appear right-skewed with similar shapes, satisfying the assumption.


# Results

```{r wilcoxon}
#wilcoxon rank-sum test (two-sided)
wilcox.test(hate_crimes_per_100k_splc ~ income_group, data = hc)
```

```{r group_medians, echo = FALSE}
#summary of median hate crime rate by income group
hc %>%
  group_by(income_group) %>%
  summarise(
    n      = n(),
    median = round(median(hate_crimes_per_100k_splc), 3),
    sd     = round(sd(hate_crimes_per_100k_splc), 3)
  )

```

The Wilcoxon rank-sum test produced a p-value of 0.027, which is less than $\alpha = 0.05$,
so the result is statistically significant. States in the above-median income group tended
to have higher hate crime rates than states in the below-median income group. The test 
statistic W reflects the number of times a hate crime rate from the above-median group ranked 
higher than one from the below-median group across all pairwise comparisons.

# Conclusions

## Interpretation

Since the p-value (0.027) is less than $\alpha = 0.05$, we reject the null hypothesis.
There is sufficient evidence to conclude that the distribution of hate crime rates
differs significantly between states with above and below median household incomes.
Interestingly, above-median income states tend to have higher hate crime rates than
below-median income states; opposite of what we might expect. This likely reflects 
the under-reporting issue noted earlier, where wealthier states may have stronger 
infrastructure for reporting hate crimes rather than genuinely experiencing more incidents.

## Limitations

Some limitations are worth noting. First, hate crime statistics are subject to significant
under-reporting, and reporting rates likely vary systematically by state wealth; this could
artificially inflate rates in wealthier states and confound our results. Second, splitting states 
into just two income groups discards information about the continuous nature of income.

## Generalizability

Since the dataset includes 46 U.S. states and D.C. (minus 4 states with missing SPLC
data), these findings describe the U.S. states included in the data as of the mid-2010s rather
than a random sample. The 4 missing states are excluded because SPLC did not record hate 
crime data for them during that time, not because of specific selections that have to do with their income
levels or hate crime rates. Their absence is unlikely to meaningfully bias the conclusions. 

That said, the results should not be generalized beyond the states included since the data are not 
based on random sampling. Still, the findings could give insight into how socioeconomic conditions 
relate to hate crime reporting across big political units, and motivate further investigation using 
individual or county-level data.

\newpage

# References

Majumder, M. (2017, January 23). *Higher rates of hate crimes are tied to income inequality.*
FiveThirtyEight. https://fivethirtyeight.com/features/higher-rates-of-hate-crimes-are-tied-to-income-inequality/

FiveThirtyEight. (2017). *Hate crimes data* [Data set]. GitHub.
https://github.com/fivethirtyeight/data/blob/master/hate-crimes/hate_crimes.csv

Federal Bureau of Investigation. (2016). *Hate crime statistics, 2015.* U.S. Department of
Justice. https://ucr.fbi.gov/hate-crime/2015

Southern Poverty Law Center. (2016). *Update: Hate incidents in the days following the election.* https://www.splcenter.org/resources/hatewatch/update-incidents-hateful-harassment-election-day-now-number-701

United States Census Bureau. (2016). *American Community Survey 5-year estimates.*
https://www.census.gov/library/visualizations/interactive/acs-5year-datamap.html

Gemini. (2026, April 19). Response to: "if i excluded 4 states out of 51 (including DC) because of missing data, should i not generalize the findings beyond the states that were included (since the data are not based on random sampling)?" https://gemini.google.com/

